---
title: "Interactive Report: Airway Data"
date: March 27, 2026
bibliography: https://gist.githubusercontent.com/johnsonra/121eb61bf4b09e4258b78e341b1819e9/raw/51d87906d0b5824f9809f667008046665b815649/references.bib
format: html
server: shiny
---

## Setup

```{r}
#| context: setup
library(shiny)
library(ggplot2)
library(dplyr)
library(scales)

# Load the airway dataset
airway_data <- read.csv("top_2000.csv", row.names = 1)

# Add a column to identify gene regulation status
airway_data <- airway_data %>%
  mutate(regulation = "Not Significant")
```

## Introduction

This interactive report provides an exploratory analysis of the airway RNA-seq dataset. Use the controls below to filter genes based on statistical significance and fold-change thresholds. The plots and statistics will update dynamically as you adjust the parameters.

## Input Controls {.sidebar}

```{r}
#| panel: sidebar
sliderInput(
  "padj_threshold",
  label = "Adjusted p-value threshold (padj):",
  min = 0.001,
  max = 0.1,
  value = 0.05,
  step = 0.001
)

numericInput(
  "lfc_threshold",
  label = "Log2 Fold-Change threshold:",
  value = 1,
  min = -5,
  max = 5,
  step = 0.1
)

hr()

textOutput("summary_text")
```

## Analysis

### Volcano Plot

```{r}
#| panel: fill
plotOutput("volcano_plot", height = 500)
```

### MA Plot

```{r}
plotOutput("ma_plot", height = 500)
```

### Summary Statistics

```{r}
tableOutput("summary_table")
```

## Reactive Expressions

```{r}
#| context: server

# Create a reactive dataset with gene annotations based on thresholds
filtered_data <- reactive({
  airway_data %>%
    mutate(
      regulation = case_when(
        padj < input$padj_threshold & log2FoldChange > input$lfc_threshold ~ "Upregulated",
        padj < input$padj_threshold & log2FoldChange < -input$lfc_threshold ~ "Downregulated",
        TRUE ~ "Not Significant"
      ),
      color = case_when(
        regulation == "Upregulated" ~ "#E74C3C",
        regulation == "Downregulated" ~ "#3498DB",
        TRUE ~ "#BDC3C7"
      )
    )
})

# Generate volcano plot
output$volcano_plot <- renderPlot({
  data <- filtered_data()
  
  ggplot(data, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
    geom_point(size = 2, alpha = 0.7) +
    scale_color_manual(
      values = c(
        "Upregulated" = "#E74C3C",
        "Downregulated" = "#3498DB",
        "Not Significant" = "#BDC3C7"
      )
    ) +
    geom_vline(xintercept = c(-input$lfc_threshold, input$lfc_threshold), 
               linetype = "dashed", color = "gray50", alpha = 0.5) +
    geom_hline(yintercept = -log10(input$padj_threshold), 
               linetype = "dashed", color = "gray50", alpha = 0.5) +
    labs(
      title = "Volcano Plot: Gene Expression Changes",
      x = "Log2 Fold-Change",
      y = "-Log10 Adjusted P-value",
      color = "Regulation"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      legend.position = "right"
    )
})

# Generate MA plot
output$ma_plot <- renderPlot({
  data <- filtered_data()
  
  # Calculate A value based on rank (since we don't have raw counts)
  data <- data %>%
    mutate(A = rank(padj))
  
  ggplot(data, aes(x = log10(A + 1), y = log2FoldChange, color = regulation)) +
    geom_point(size = 2, alpha = 0.7) +
    scale_color_manual(
      values = c(
        "Upregulated" = "#E74C3C",
        "Downregulated" = "#3498DB",
        "Not Significant" = "#BDC3C7"
      )
    ) +
    geom_hline(yintercept = c(-input$lfc_threshold, input$lfc_threshold), 
               linetype = "dashed", color = "gray50", alpha = 0.5) +
    geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha = 0.3) +
    labs(
      title = "MA Plot: Expression Changes",
      x = "Log10(Gene Rank)",
      y = "Log2 Fold-Change (M)",
      color = "Regulation"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      legend.position = "right"
    )
})

# Generate summary statistics
output$summary_table <- renderTable({
  data <- filtered_data()
  
  summary_stats <- data %>%
    group_by(regulation) %>%
    summarise(
      `Number of Genes` = n(),
      `Mean Log2FC` = round(mean(log2FoldChange), 3),
      `Median Log2FC` = round(median(log2FoldChange), 3),
      `Mean -Log10(padj)` = round(mean(-log10(padj)), 3),
      .groups = 'drop'
    ) %>%
    arrange(factor(regulation, levels = c("Upregulated", "Downregulated", "Not Significant")))
  
  summary_stats
}, striped = TRUE, bordered = TRUE)

# Generate summary text
output$summary_text <- renderText({
  data <- filtered_data()
  
  upregulated <- sum(data$regulation == "Upregulated")
  downregulated <- sum(data$regulation == "Downregulated")
  not_sig <- sum(data$regulation == "Not Significant")
  total <- nrow(data)
  
  paste0(
    "Summary:\n\n",
    "Upregulated: ", upregulated, "\n",
    "Downregulated: ", downregulated, "\n",
    "Not Significant: ", not_sig, "\n\n",
    "Total Genes: ", total
  )
})
```

## References
