---
title: "Airway Study: RNA-seq Analysis Report"
date: March 27, 2026
bibliography: https://gist.githubusercontent.com/johnsonra/121eb61bf4b09e4258b78e341b1819e9/raw/51d87906d0b5824f9809f667008046665b815649/references.bib
format: gfm
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(scales)

# Load the top 2000 DEGs data
degs <- read.csv("top_2000.csv", row.names = 1)

# Convert p-values for visualization
degs$neg_log10_padj <- -log10(degs$padj)
```

# Airway Study: RNA-seq Transcriptome Analysis

This report presents a differential gene expression analysis of the airway dataset from @himes_rna-seq_2014. The study examines transcriptomic changes in airway smooth muscle cells in response to dexamethasone treatment using RNA-seq technology.

## Interactive Version

Want to explore the data interactively? Check out the **[Interactive Shiny Report](Airway_shiny.html)** where you can adjust significance thresholds and dynamically filter the genes in real-time!

## Volcano Plot of Differentially Expressed Genes

The volcano plot below displays the log2 fold-change versus the adjusted p-value for the top 2,000 most significant genes. Points are colored by direction of expression change: upregulated genes (positive log2FC) are shown in red, while downregulated genes (negative log2FC) are shown in blue.

```{r}
#| fig-width: 10
#| fig-height: 7
#| echo: false

# Create volcano plot
volcano_data <- degs %>%
  mutate(
    significant = padj < 0.05,
    direction = ifelse(log2FoldChange > 0, "Upregulated", "Downregulated")
  )

ggplot(volcano_data, aes(x = log2FoldChange, y = neg_log10_padj)) +
  geom_point(aes(color = direction), alpha = 0.6, size = 3) +
  scale_color_manual(values = c("Upregulated" = "#e74c3c", "Downregulated" = "#3498db")) +
  labs(
    title = "Volcano Plot: Differentially Expressed Genes",
    subtitle = "Airway study - dexamethasone treatment vs. untreated control",
    x = "log2(Fold Change)",
    y = "-log10(adjusted p-value)",
    color = "Expression Change"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray50"),
    axis.title = element_text(size = 12),
    legend.position = "right",
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank()
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray70", alpha = 0.5)
```

## MA Plot: Expression vs. Fold Change

The MA plot presents the average log2 expression level (x-axis) versus the log2 fold-change (y-axis) for all 2,000 genes. This visualization helps identify expression patterns across different abundance levels.

```{r}
#| fig-width: 10
#| fig-height: 7
#| echo: false

# Create MA plot data (using rank as proxy for expression level)
ma_data <- degs %>%
  mutate(
    rank_order = seq_len(n()),
    A = log2(rank_order + 1),  # Rank-based pseudo expression
    M = log2FoldChange,         # Log fold change
    direction = ifelse(M > 0, "Upregulated", "Downregulated")
  ) %>%
  filter(!is.na(M))

ggplot(ma_data, aes(x = A, y = M)) +
  geom_point(aes(color = direction), alpha = 0.5, size = 2) +
  scale_color_manual(values = c("Upregulated" = "#e74c3c", "Downregulated" = "#3498db")) +
  labs(
    title = "MA Plot: Expression Level vs. Log2 Fold-Change",
    subtitle = "Red points: upregulated | Blue points: downregulated",
    x = "A: log2(Average Expression)",
    y = "M: log2(Fold Change)",
    color = "Expression Change"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray50"),
    axis.title = element_text(size = 12),
    legend.position = "right",
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank()
  ) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray70", alpha = 0.5)
```

## Summary Statistics

```{r}
#| echo: false

# Calculate summary statistics
upregulated <- sum(degs$log2FoldChange > 0 & degs$padj < 0.05, na.rm = TRUE)
downregulated <- sum(degs$log2FoldChange < 0 & degs$padj < 0.05, na.rm = TRUE)
total_significant <- upregulated + downregulated

cat(sprintf("**Total genes analyzed:** %d\n\n", nrow(degs)))
cat(sprintf("**Significantly regulated genes (padj < 0.05):** %d\n\n", total_significant))
cat(sprintf("  - Upregulated: %d\n\n", upregulated))
cat(sprintf("  - Downregulated: %d\n\n", downregulated))

# Top 10 most significant genes
top_10 <- degs %>%
  arrange(padj) %>%
  head(10) %>%
  select(symbol, log2FoldChange, pvalue, padj)

cat("**Top 10 Most Significant Genes:**\n\n")
print(top_10)
```

## References

This analysis is based on the RNA-seq dataset and methods described in @himes_rna-seq_2014.
